%%
%% This is a LaTeX document formatted using the ACM 'acmlarge' style.
%% It presents a scientific report on User Control in Tourism Recommender Systems.
%%
\documentclass[acmlarge]{acmart}

\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

\usepackage{graphicx}

%% Rights management information.
\setcopyright{none}
\settopmatter{printacmref=false} % Removes ACM Reference Format
\acmJournal{POMACS}
\acmVolume{1}
\acmNumber{1}
\acmArticle{1}
\acmYear{2026}
\acmMonth{6}
\acmDOI{}

\begin{document}

\title[Empowering Users with Algorithmic Control and Transparency]{Empowering Users with Algorithmic Control and Transparency in Tourism Recommender Systems: A DSA-Compliant Approach}

\author{David Gallob}
\email{david.gallob@tum.de}
\affiliation{%
  \institution{Technical University of Munich}
  \department{School of Computation, Information and Technology}
  \city{Munich}
  \country{Germany}
}

\renewcommand{\shortauthors}{Gallob}

\begin{abstract}
The recommender systems people encounter in daily life are very often "black boxes", which means that users rarely know how their recommendations are generated and they have no control over which items are recommended. In tourism, this leads to a lack of trust and user satisfaction, as users can never be sure if their recommendation is really optimal or influenced by profit maximization goals of the platform. The EU has therefore introduced the Digital Services Act (DSA), a framework which requires platforms to explain their algorithms in an understandable way and offer users a profiling-free alternative. In this paper a web-based prototype for hotel recommendations which addresses these issues by implementing two algorithms is presented. The web-app offers detailed explanations of the underlying decision algorithms and shows for each result, how the recommendation was computed. The two algorithms which come to use are Weighted Multi-Attribute Utility Theory (MAUT), which lets users set weights on the different attributes (e.g. stars, distance to center, price) and Constraint Filtering, where the user sets hard limits to the attributes (e.g. price must be below 100 Euro per night). This paper details the system architecture as well as the mathemical foundations of the used algorithms along with the UI features explaining the algorithms to the user
\end{abstract}

\keywords{Recommender Systems, User Control, Multi-Attribute Utility Theory, Constraint Filtering, Digital Services Act, Explainable AI, E-Tourism}

\maketitle

\section{Introduction}
In the day-to-day life, recommender systems (RS) are ubiquitous from online shopping and video streaming to hotel and flight recommendation plaftorms~\cite{adomavicius2005toward}. Many prominent RS do not let users know how their recommendations are generated and offer no direct means of influencing the outcome~\cite{BORRAS20147370}. Those systems are often built on complex algorithms (mainly neural networks) which process enormous amounts of user to generate suggestions. They rarely explain users their underlying logic. This lack of transparency can hinder trust and user satisfaction and leads to frustration and cognitive overhead. The tourism domain where a high financial expenditure and the emotional investment is generally high is particularly affected by this issue~\cite{ricci2011travel}.

The Human-Computer Interaction (HCI) field has focused their research on user influence and transparency (explainability) to address these shortcomings~\cite{knijnenburg2012explaining, tintarev2012evaluating}. Instead of using automated profiling, interactive recommender systems give users the ability to influence the recommendation process by adjusting sliders which represent their importance or setting hard constraints for the attributes (e.g., price or distance).

The demand for transparency is further fueled by the European Union's Digital Services Act (DSA), which governs online platforms and aims to protect user's rights. The DSA mandates that online platforms must provide users with transparent information about the main parameters of their recommender systems in an understandable language in plain text and offer options to influence them (Aritcle 27)~\cite{eudsa2022}. Article 38 further obliges very large online platforms (VLOPs) to provide users with at least one option for their recommender systems that is not based on profiling. These articles require developers to shift from black-box profiling towards transparant and user controlled platforms.

In this paper the design decisions, architecture and implementation of a web-application tourism recommender system prototype which is compliant with the DSA is detailed. It adresses the transparency as well as the user influence goals by implementing two algorithms:
\begin{enumerate}
    \item \textbf{Algorithm A: Weighted Multi-Attribute Utility Theory (MAUT)}:
    This algorithm uses sliders to determine weights for hotel attributes (price, location, star rating, customer rating, and public transit access). The algorithm performs a normalization of the attributes and computes a weighted utility score for each hotel. Finally the hotels are ranked according to their utility score (percentage match).
    \item \textbf{Algorithm B: Constraint-Based Filtering}: 
    In Constraint-Based Filtering the user defines hard limits for each attribute (e.g. maximum price or minimum rating). All hotels which do not meet the contraints are filtered out. This allows users to specify their preferences in a clear and understandable way without the need for complex mathematical calculations.
\end{enumerate}

The prototype features a banner at the top of the landing page explaining the chosen algorithm in clear words and offers a button which when pressed shows the mathematical formula of the chosen algorithm for enhanced transparency. The calculated results also offer a collapsible mathematical breakdown (only on user intent to not overwhelm not interested users) showing a step-by-step calculation how the result's score was computed. This allows users to understand how the results are generated and how their inputs influence the outcome. The calculation breakdown is generated using KaTeX and renders the exact mathematical formulas and normalization equations used in the calculations, allowing users to trace how their inputs directly translate into the presented results.


The remainder holds the following structure: In \autoref{sec:related_work} 
the related work in current tourism recommender systems, MAUT and explaination interfaces is identified. \autoref{sec:system_architecture_and_implementation} details the system's architecture and the mock dataset. \autoref{sec:algorithmic_framework} shows the mathematical foundations of the two used algorithms. \autoref{sec:user_interface_design_and_explainability} details the UI design and transparency features. \autoref{sec:future_work} proposes a protocol for a future user study and extensions, and \autoref{sec:conclusion} concludes the paper. In the Appendix the complete mock database used in the project, followed by the author contribution table and GenAI disclosures are attached.
%cont here
\section{Related Work}\label{sec:related_work}
Recommender systems' development face special challenges which are unique to their domain --- such as recommendations for movies and and music~\cite{BORRAS20147370}. In the toursim domain decision require a lot of rational resources, are infrequent and have multiple decision attributes. In contrast to to low-cost consumer goods, vacation planning involves a high final risk and high emotional expectations, which also leads to a high time commitment. Users therefore take extensive information and comparisons into account and need to express specific constraints before making a final decision instead.~\cite{ricci2011travel}.

Traditional recommender systems (TRSs) in the tourism domain are mainly divided into 3 categories~\cite{Sarkar2023, 8308293}:
\begin{itemize}
  \item \textbf{Collaborative filtering:} Builds on historical information of similar users, but suffers from the Cold Start Problem --- New hotels without reviews and seldom travellers without historical ratings lead to wrong recommendations
  \item \textbf{Content-based:} These systems match item features with the preference profiles of the user. While they get good results, they are often unable to explain their recommendations and to adapt to rapid changes in user needs (e.g., business travelers in central locations versus family budget accommodations)
  \item \textbf{Hybrid:} These recommender systems combine multiple approaches.
\end{itemize}

In order to tackle these limitations, multi-attribute frameworks,i.e. Multi-Attribute Utility Theory (MAUT) for decision making have been coming to use interactive recommender systems~\cite{schmitt2002maut}. MAUT is a mathematical algorithm which allows to evaluate alternatives with potentially conflicting attributes\cite{keeney1993decisions}. The algorithm defines a utility function for every attribute (e.g., mapping location to a utility between 0 and 1) and calculates a weighted sum. MAUT, as shown bei Schmitt et al. in 2002~\cite{schmitt2002maut}, is particularly useful in interactive recommender systems, as it takes huma decision making processes into account (i.e., user defined weights). It allows for deficiencies (e.g. high price) to be evened out by a high value in another attribute (e.g., excellent location).

Constraint-based recommender systems are non compensatory in contrast. They do not calculate a score, but filter out items violating user-defined limits based on explicit knowledge about the items~\cite{ricci2011travel}. This approach is easy to understand and does not require user profiling, perferctly aligning with the EU's DSA~\cite{eudsa2022}.

Earlier research has shown that the success of interactive recommender systems strongly relies on their UIs~\cite{mcnee2006being,swearingen2001beyond}. Therefore providing explanation in the UI improves the user understanding as well as their trust and their perceived control over the system~\cite{tintarev2012evaluating}. Knijnenburg et al.~\cite{knijnenburg2012explaining} showed that their proposed framework offering user control paired with high explainability results in higher user satisfaction and acceptance. The integration of sliders for defining weights and mathematical breakdowns along with transparency panels investigate in this paper builds on those findings to build an explicable and transparent UI for user satisfaction and to meet the DSA requirements.

\section{System Architecture and Implementation}\label{sec:system_architecture_and_implementation}
The prototype is designed as a client-side, responsive single-page application (SPA). The technical stack consists of React\footnote{\url{https://react.dev}}, Vite\footnote{\url{https://vite.dev}} for the build system, and vanilla CSS for layout and styling. Client-side math rendering is powered by KaTeX\footnote{\url{https://katex.org}}, a fast LaTeX math rendering library, integrated via a custom React wrapper.The system architecture is structured to support real-time user interaction.

The application state manages:
\begin{itemize}
  \item The selected city (Munich vs. Vienna).
  \item The active algorithmic mode (MAUT vs. Constraint based).
  \item The five attribute weights ($w_i \in [0, 100]$) for the MAUT mode.
  \item The five attribute limits for the Constraint Filtering mode.
  \item The global formula toggle (controlling whether plain-language explanations or mathematical LaTeX formulas are shown in the transparency banners).
\end{itemize}

The dataset, stored in a static module (`data.js`), comprises 20 hotels (10 in Vienna and 10 in Munich). The hotels were designed to represent a realistic and diverse spectrum of accommodations, ranging from budget-friendly youth hostels to luxury five-star resorts. Each hotel is defined by a unique identifier, name, city, description, and five key quantitative attributes:
\begin{enumerate}
  \item \textbf{Price per Night }: Ranging from \$35 (Hostel Vienna West) to \$550 (Mandarin Oriental Munich).
  \item \textbf{Distance to Center}: Distance from the historical city center (Stephansplatz in Vienna, Marienplatz in Munich). Ranging from 0.1 km (Stephansplatz) to 28.0 km (Airport Express Inn Munich).
  \item \textbf{Star Rating}: Ranging from 1 (basic tourist class) to 5 stars (luxury class).
  \item \textbf{Guest Review Score}: A continuous rating between 1.0 (worst) and 10.0 (excellent), representing aggregated guest satisfaction.
  \item \textbf{Public Transport Access Score}: A rating between 1.0 (poor) and 10.0 (excellent), representing proximity to subway stations, tram lines, and frequency of transit services.
\end{enumerate}

The application recalculates recommendations in real time whenever the user modifies a slider or toggles a control. Because the calculations are performed client-side on a bounded dataset, updates occur instantly, providing immediate visual feedback.

\section{Algorithmic Framework}\label{sec:algorithmic_framework}
To support the user-controlled interaction model, we formalize the mathematical models behind the two algorithmic recommendation modes.

\subsection{Algorithm A: Weighted Multi-Attribute Utility Theory (MAUT)}
Weighted MAUT evaluates each hotel $X_j$ in the active city set $H_{\text{city}}$ by mapping its raw attribute values to normalized utility scores and computing a weighted average. The calculation proceeds in three stages:

\subsubsection{Normalization of Attributes}
Since attributes are measured in different units (USD, kilometers, star counts, and continuous scales), they must be normalized to a standard utility scale $u_i(x) \in [0, 1]$, where $0$ represents the least desirable outcome and $1$ represents the most desirable.

The attributes are categorized into two types: cost metrics (where lower values are better) and benefit metrics (where higher values are better).
Furthermore, the system employs two different normalization boundaries:
\begin{itemize}
  \item \textbf{Relative Normalization}: Normalizes values based on the maximum and minimum values present within the active city's dataset, maximizing the contrast between local options.
  \item \textbf{Absolute Normalization}: Normalizes values based on a fixed, theoretically defined range (e.g., 1 to 5 stars, or 1 to 10 rating scale).
\end{itemize}

\subsubsection{Weighted Aggregation}
Let $W = (w_{\text{price}}, w_{\text{location}}, w_{\text{stars}}, w_{\text{rating}}, w_{\text{transport}})$ be the vector of weights specified by the user, where each weight $w_i \in [0, 100]$. The total utility $U(X_j)$ for hotel $X_j$ is calculated as:
\begin{equation}
  U(X_j) = \frac{\sum_{i \in \text{Attrs}} w_i \cdot u_i(x_{ij})}{\sum_{i \in \text{Attrs}} w_i} \times 100\%
\end{equation}
where $\text{Attrs} = \{\text{price}, \text{location}, \text{stars}, \text{rating}, \text{transport}\}$ and $x_{ij}$ represents the raw value of attribute $i$ for hotel $j$. The resulting utility $U(X_j) \in [0, 100]\%$ represents the percentage match score. Hotels are sorted and presented to the user in descending order of $U(X_j)$.

\subsection{Algorithm B: Constraint-Based Filtering}
Constraint-Based Filtering does not calculate a numerical match score or rank hotels by preference. Instead, it screens the active city's dataset against five user-defined limits. A hotel is included in the recommended set if and only if it satisfies the conjunction of all active constraints. This represents a non-compensatory filtering logic. A hotel that violates even a single constraint (such as exceeding the maximum price by \$5) is immediately excluded from the results, regardless of whether it possesses perfect scores (e.g., 5 stars, 10.0 user rating) in all other categories.

\subsection{Comparison of Decision Logic}
The two models represent fundamentally different decision-making styles, as summarized in Table~\ref{tab:comparison}.

\begin{table}[h]
  \caption{Comparison of the Implemented Algorithmic Paradigms}
  \label{tab:comparison}
  \begin{tabular}{lll}
    \toprule
    Dimension & Algorithm A: Weighted MAUT & Algorithm B: Constraint Filtering \\
    \midrule
    \textbf{Decision Logic} & Compensatory (trade-offs allowed) & Non-compensatory (hard boundaries) \\
    \textbf{Output Type} & Ranked list with continuous match scores & Unordered subset of matching items \\
    \textbf{User Controls} & Relative importance weights ($w_i \in [0, 100]$) & Threshold limits ($P_{\text{max}}, D_{\text{max}}$, etc.) \\
    \textbf{Cognitive Style} & Optimizing & Screening / Satisficing \\
    \textbf{DSA Compliance} & Explains weights & Explains inclusion/exclusion criteria \\
    \textbf{Profiling Dependency} & None (Profiling-free, real-time input) & None (Profiling-free, real-time input) \\
    \bottomrule
  \end{tabular}
\end{table}

\section{User Interface Design and Explainability}\label{sec:user_interface_design_and_explainability}

The prototype interface was designed with a glassmorphism aesthetic (utilizing translucent backdrops, vibrant background gradients, and smooth animations) to create a premium, modern user experience. More importantly, the interface was structured to maximize explainability and user control.

\subsection{DSA-Compliant Explanation Interfaces}
To comply with the requirements of the Digital Services Act, transparency mechanisms are built directly into the primary interaction flow rather than being hidden in auxiliary pages.

\subsubsection{Global Explanation Banners}
At the top of the interface, a dedicated, colored information card dynamically adapts to the selected algorithm.
\begin{itemize}
  \item In \textbf{Weighted MAUT} mode, the banner displays a blue theme, introducing the concept of multi-attribute weighting.
  \item In \textbf{Constraint-Guided} mode, the banner displays a green theme, introducing the logic of set reduction and non-compensatory filtering.
\end{itemize}

Both banners include a "Show Mathematical Details" toggle button. Clicking this button expands a sub-panel that renders the mathematical equations in LaTeX (Equation 6 for MAUT, and Equation 9 for Constraint-Guided Filtering) alongside a list of definitions. This design provides a tiered explanation model: novice users receive a brief, plain-language summary of how the parameters are aggregated, while expert users or regulators can inspect the exact mathematical formulas.

\subsubsection{On-Demand Collapsible Math Breakdowns}
For the Weighted MAUT mode, each hotel card features a "Show Mathematical Breakdown" button. Expanding this panel reveals a step-by-step trace of the utility calculation for that specific hotel.
The panel displays:
\begin{itemize}
  \item The general formula
  \item A tabular breakdown for each of the five attributes, displaying the normalized utility score, the user-defined weight, and the product.
  \item The summation step, showing the sum of products divided by the sum of weights.
  \item The final rounded match percentage.
\end{itemize}

This feature provides absolute transparency, eliminating the perception of the recommender system as a black box. The user can see exactly why a hotel is ranked first or second, and can directly trace the mathematical impact of adjusting a slider.


\section{Future Work}\label{sec:future_work}
While the current prototype demonstrates the feasibility of user-controlled, transparent tourism recommenders, several research and development paths remain open.

\subsection{The Hybrid Algorithmic Approach}
A natural evolution of the two modes is a hybrid algorithm that combines the strengths of both. Currently:
\begin{itemize}
  \item Weighted MAUT is compensatory, which can lead to high cognitive load as users must constantly balance multiple sliders to filter out unacceptable options.
  \item Constraint-Based Filtering is non-compensatory and does not rank the matching items, which can leave the user with too many options or an empty set.
\end{itemize}
A hybrid recommender would operate as a two-stage pipeline:
\begin{enumerate}
  \item \textbf{Stage 1: Constraint Filtering (Non-Compensatory screening)}: The system applies user-defined constraints to filter out hotels that are completely unacceptable (e.g., price $> \$250$, or distance $> 10$ km). This reduces the candidate pool to a set of feasible options.
  \item \textbf{Stage 2: Weighted MAUT (Compensatory ranking)}: The system applies the user's preference weights to rank only the hotels that passed Stage 1.
\end{enumerate}
This hybrid approach represents a realistic model of human decision-making, where decision-makers first eliminate unacceptable alternatives and then optimize among the survivors. It maintains full transparency and complies with the DSA, while reducing user cognitive load.

\subsection{Proposed User Evaluation Protocol}
To evaluate the user experience (UX) and the effectiveness of the explainability features, we propose a controlled user study protocol.

\subsubsection{Hypotheses}
We formulate three primary hypotheses:
\begin{itemize}
  \item \textbf{H1 (Perceived Control)}: Users interacting with the Weighted MAUT or Constraint Filtering interfaces report a higher sense of control over recommendations compared to a baseline "black-box" interface.
  \item \textbf{H2 (System Trust)}: Exposing the collapsible mathematical breakdown panels increases user trust in the system's recommendations without causing information overload.
  \textbf{H3 (DSA Compliance Impact)}: The presence of the global explanation banner and KaTeX formulas increases user comprehension of the algorithm's inner workings.
\end{itemize}

\subsubsection{Experimental Design}
We propose a within-subjects laboratory study with $N=30$ participants. Each participant will perform travel planning tasks under three experimental conditions:
\begin{enumerate}
    \item \textbf{Condition 1 (Baseline - Black Box)}: A traditional recommender interface displaying hotel cards ranked by a hidden algorithm, with no sliders and no explanation panels.
    \item \textbf{Condition 2 (Weighted MAUT with On-Demand Explanations)}: The interactive slider-based interface with collapsible math breakdown panels.
    \item \textbf{Condition 3 (Constraint Filtering with Explanations)}: The constraint-based interface showing absolute threshold sliders.
\end{enumerate}

The ordering of conditions will be counterbalanced to prevent learning effects. Following each condition, participants will complete standardized questionnaires to measure:
\begin{itemize}
  \item \textbf{System Usability}: Measured via the System Usability Scale (SUS).
  \item \textbf{User Experience \& Trust}: Measured using the ResQue (Recommender Systems Quality of Experience) framework, focusing on trust, perceived accuracy, control, and transparency.
  \item \textbf{Cognitive Load}: Measured via the NASA-TLX (Task Load Index) to evaluate if exposing mathematical details increases mental demand.
\end{itemize}
The results of this study will guide the refinement of the UI, helping to strike a balance between complete mathematical transparency and user cognitive ease.

\section{Conclusion}\label{sec:conclusion}
The era of black-box recommender systems is facing both user-experience and regulatory challenges. This paper presented an interactive, transparent tourism recommender system prototype designed to address the compliance requirements of the EU Digital Services Act (DSA). By implementing both a compensatory Weighted Multi-Attribute Utility Theory (MAUT) model and a non-compensatory Constraint-Based Filtering model, the prototype provides users with direct, profiling-free control over recommendation logic. We integrated tiered explainability features, combining simple plain-language summaries with on-demand, KaTeX-rendered mathematical breakdowns. Finally, we analyzed and resolved mathematical edge cases, including division-by-zero singularities and score overshooting, ensuring a stable and robust system. This architecture serves as a blueprint for developers seeking to combine rich user control, modern web aesthetics, and strict regulatory compliance in recommender system design.

\begin{acks}
The author would like to thank the teaching staff of the Lab Course: Projects in Recommender Systems (SS 2026) at the Chair of Connected Mobility, Technical University of Munich, for their guidance and feedback throughout the project.
\end{acks}

\bibliographystyle{ACM-Reference-Format}
\bibliography{sample-base}

\appendix

\section{Complete Hotel Database}
Table~\ref{tab:database} presents the complete dataset of 20 hotels used in the recommender system prototype, split between Vienna (v1--v10) and Munich (m1--m10).

\begin{table}[h]
  \caption{Hotel Database Attributes}
  \label{tab:database}
  \begin{tabular}{llrrrrl}
    \toprule
    ID & Hotel Name & Price (\$/night) & Distance (km) & Stars & Rating (/10) & Transport (/10) \\
    \midrule
    \multicolumn{7}{c}{\textbf{Vienna Hotels}} \\
    \midrule
    v1 & Grand Imperial Vienna & 350 & 0.5 & 5 & 9.6 & 9.5 \\
    v2 & Boutique Hotel am Stephansplatz & 180 & 0.1 & 4 & 9.2 & 10.0 \\
    v3 & Danube Riverside Hotel & 95 & 4.5 & 3 & 8.0 & 6.5 \\
    v4 & Schönbrunn Palace Suites & 290 & 6.0 & 5 & 9.8 & 8.0 \\
    v5 & Hostel Vienna West & 35 & 3.2 & 1 & 7.5 & 9.0 \\
    v6 & Art Deco Hotel Prater & 140 & 2.8 & 4 & 8.7 & 8.5 \\
    v7 & Economy Inn Vienna & 60 & 5.5 & 2 & 6.8 & 5.0 \\
    v8 & Museum Quarter Retreat & 210 & 1.2 & 4 & 9.4 & 9.0 \\
    v9 & Green Oasis Hotel & 110 & 7.5 & 3 & 8.9 & 4.0 \\
    v10 & The Ritz-Carlton Vienna & 450 & 0.8 & 5 & 9.7 & 10.0 \\
    \midrule
    \multicolumn{7}{c}{\textbf{Munich Hotels}} \\
    \midrule
    m1 & Bayerischer Hof & 420 & 0.4 & 5 & 9.5 & 9.5 \\
    m2 & Marienplatz Business Hotel & 160 & 0.2 & 4 & 8.8 & 10.0 \\
    m3 & Oktoberfest Lodge & 85 & 2.5 & 2 & 7.2 & 8.0 \\
    m4 & Englischer Garten Retreat & 250 & 3.0 & 4 & 9.3 & 7.0 \\
    m5 & Schwabing Youth Hostel & 40 & 4.0 & 1 & 8.1 & 8.5 \\
    m6 & Olympic Park Hotel & 130 & 6.5 & 3 & 8.5 & 7.5 \\
    m7 & Airport Express Inn & 90 & 28.0 & 3 & 7.8 & 9.0 \\
    m8 & Viktualienmarkt Boutique & 190 & 0.6 & 4 & 9.1 & 9.5 \\
    m9 & Isar River View & 210 & 1.8 & 4 & 8.9 & 6.0 \\
    m10 & Mandarin Oriental Munich & 550 & 0.5 & 5 & 9.8 & 9.5 \\
    \bottomrule
  \end{tabular}
\end{table}

\section{Author Information and Contribution Details}
The author, David Gallob (david.gallob@tum.de) worked alone in this project.

\section{Generative AI Disclosure}
This project utilized Generative AI tools (specifically Google DeepMind's Gemini 3.5 Flash) during development. The AI was employed as a pair-programming assistant to:
\begin{itemize}
  \item Draft and optimize the user interface layouts and CSS styling rules.
  \item Debugging and recommending JavaScript state handling best practices.
  \item Assist in structuring, polishing and editing the LaTeX code of this scientific report.
\end{itemize}
All design decisions, algorithmic formulations, bug resolutions, and report text were audited, verified, and approved by the human author to ensure compliance with the scientific standards of the Technical University of Munich.
\end{document}
\endinput
